Coordinated monitoring of the Piney Point wastewater discharge into Tampa Bay: Data synthesis and reporting

### 2022-05-20 manu-draft.Rmd
discussion draft done

* Added value of the app
    * analytics speak for themselves, but not the whole picture.  How did users get to our dashboard (some of this info is in Analytics, but which of our comms products drove the bus)? How was it used (e.g., screenshots for pres, K. Hubbard pers. comm.)?
    * how can we better assess the value of these tools more holistically? What is the true measure of success in terms of behavior change or driving policy, if any? 
* Relevant considerations
    * What worked - use of OS tools, but still very time-consuming (e.g., R/dat_proc.R is thousands of lines of code)
    * Need for unified datasets, use of tidy principles, etc. Also mention that this relied on partners sending us data/putting data on GDrive, not everyone did that (note issues with the latter, i.e., access from elsewhere can be unreliable, but we also didn't have all the data).  Highly value of partnerships - data collection and sharing.
    * Value of historical data and contextualized summaries
    * Exposure of additional data needs (e.g., macroalgae)
* Future of Piney Point, next steps, etc.

>>>>

The release of wastewater from Piney Point in the spring of 2021 was an unfortunate event that mobilized dozens of local resource managers and stakeholders to track the potential impacts to Tampa Bay. The communication tools and workflows used to synthesize the data collected in this effort were useful for informing the management response and the analytics of our dashboard suggested that the products created by TBEP were valuable resources in 2021.  However, dashboard analytics are not the only metric of success and additional information can help assess the effectiveness of these tools to efficiently and transparently communicate science for decision-making.  Area researchers and managers were able to use screenshots from the dashboard to communicate results in real time, although there was no formal metric for these applications.  More holistically assessing the value of open science tools is also needed, such as how they may be used to inform policy or regulatory changes aimed at minimizing environmental impacts.

From a data science perspective, several lessons were learned from the need to rapidly synthesize multiple sources of information collected during the response-based monitoring.  Broader use of "tidy" data principles would significantly decrease the processing time required to create data in a usable format for analysis.  Data entry sheets or digital platforms for sharing data should be structured to more easily transform the the results into tidy format (i.e., one observation per row, one varialbe in a column, one value per cell).  However, this requires coordination among multiple entities as to which standards are acceptable for use in digital information storage and even additional training activities to bridge communication gaps between field staff and analysts. Data wrangling will always be a component of analysis, but the time spent doing so could be reduced by communication between institutions to adopt similar data standards.  Finding solutions for sharing data was also problematic.  Although Google Drive was useful, not everyone was able to access the information given IT limitations that were agency-specific.  Moreover, hosting data on Google Drive was contingent on trust and willingness to share data with TBEP.  The TBEP did not have access to all data collected for Piney Point, despite the dozens of partners that did share data with the organization.  Building and sustaining partnerships is a fundamental mission of the National Estuary Program.  Past successes of TBEP have helped build this trust and the continued success of the organization is dependent on sustaining and growing partnerships in the future.

The permanent closure of Piney Point in an environmentally responsible manner remains the ultimate goal of regional resource managers and policy-makers.  The open science tools created by the TBEP in 2021 were effective based on the limited means of assessing their impact, yet it is hoped that the tools can facilitate conversations towards broader goals.  In addition to Piney Point, mining activities in Florida and elsewhere continue to impose external costs on human health and the environment. The role of science in documenting these impacts is important, but even more so, research priorities should focus on understanding societal and political mechanisms that enable these preventable impacts to occur.  Combining traditional science with open science technologies can effectively enable positive change through better science communication.
